---
title: "Multiple Systems Estimation Example"
output:
  html_document:
    code_folding: hide
date: "2024-08-01"
---

Based on the methods proposed by Lax Chan, Bernard W. Silverman & Kyle Vincent in Multiple Systems Estimation for Sparse Capture Data: Inferential Challenges When There Are Nonoverlapping Lists, Journal of the American Statistical Association, 116:535, 1297-1306, DOI: 10.1080/01621459.2019.1708748
# https://rdrr.io/cran/SparseMSE/man/estimatepopulation.0.html


This document illustrates an approach for organizing administrative data to utilize a method for estimating the true size of a population when that is difficult to observe directly, called Multiple Systems Estimation (MSE). In order to take advantage of MSE, it is necessary to have information about the existence of at least some members of the target population recorded in multiple data sources, and for the observations made in one data source to be linkable to the other data sources, such that it is possible to determine whether a person recorded in data source A is only recorded in A, or if they are also recorded in some combination of B, C, and D. 

To demonstrate this approach, we created an example synthetic dataset, _mse_synth_data.csv_, of youth facing homelessness in Cuyahoga County, Ohio in 2017, 2018, and 2019, as recorded in four linked administrative data sources (named list1, list2, list3, and list4). It includes a series of binary flags that indicate if the individual was identified in one of four administrative databases  as having experienced homelessness or housing instability in a given year (e.g., list1_17). 


Each record also includes basic personal characteristics used to stratify youth into homogenous subgroups along the lines of race (white, non-white), sex assigned at birth (M, F), and age group (13-18, 19-25). A date of birth was randomly assigned to each record based on age group. 


This code accomplishes three things. First, it organizes and reshapes a linked administrative dataset into the format needed to carry out MSE.  Second, it demonstrates an efficient method for carrying out the same set of analyses on numerous subsets of the data simultaneously, and without excessive code repetition. Third, it extracts some key pieces of information from the MSE output to allow researchers to begin drawing conclusions about the population of homeless youth in Cuyahoga County, the degree to which the homeless population is 'hidden' and thus unlikely to be able to be reached by service providers, and for better understanding differences between different subpopulations of homeless youth, along the lines of race, sex, and age.  

The table below shows the initial data structure.



```{r setup, include=FALSE}

#loading R Packages/libraries & reading data

library(reactable)
library(SparseMSE)
library(tidyverse)

dat <- read.csv("mse_synthdata.csv")

dat <- dat %>%
  mutate(dob = as_date(dob))


```

```{r echo=FALSE, message=FALSE, warning=TRUE}
reactable(slice_sample(dat, n = 10), fullWidth = TRUE)

```




To begin, we identify and retain the records for individuals observed in at least one list in a year in which they were between the ages of 13 and 25. We discard records for individuals who never appear on any list, or who only appear on a list when they are outside the age range for this study (under age 13; age 26 or older)


```{r message=FALSE, warning=FALSE}
dat2 <- dat %>% 
  mutate(
    elig17 = ifelse(dob %within% interval(as_date("1992-01-01"), as_date("2004-12-31")), 1, 0), 
    elig18 = ifelse(dob %within% interval(as_date("1993-01-01"), as_date("2005-12-31")), 1, 0), 
    elig19 = ifelse(dob %within% interval(as_date("1994-01-01"), as_date("2006-12-31")), 1, 0), 
    observed17 = ifelse(list1_17 + list2_17 + list3_17 + list4_17 > 0, 1, 0), 
    observed18 = ifelse(list1_18 + list2_18 + list3_18 + list4_18 > 0, 1, 0), 
    observed19 = ifelse(list1_19 + list2_19 + list3_19 + list4_19 > 0, 1, 0), 
    # id = row_number()
  )


```


From the initial pool of records, we identified those for youth who were recorded as facing homelessness in a year in which they were in the age range for this study.  


```{r}

dat3 <- dat2 %>% 
  filter(
    (elig17 == 1 & observed17 == 1) |
    (elig18 == 1 & observed18 == 1) |
    (elig19 == 1 & observed19 == 1) 
  )


```



Next, we reshape the dataset from "wide" (all information about a person contained in a single row of data, with more columns) to "long" (with one row for each administrative list and year in which an individual was recorded, such that a person recorded as facing homelessness in all four lists every year from 2017-2019 span 12 data rows). This is done to make it easier to retain only information about people in years in which they were recorded. 



```{r}

dat_elig <- dat3 %>% 
  select(id, starts_with("elig")) %>% 
  pivot_longer(cols = -id) %>% 
  mutate(source = str_sub(name, 1, -3),list_yr = as.numeric(paste0("20", str_sub(name, -2)))) %>% 
  rename(eligible = value) %>% 
  select(id, list_yr, eligible) %>% 
  distinct()


dat_long <- dat3 %>% 
  select(-starts_with("elig")) %>% 
  pivot_longer(cols = c(starts_with("list1_"), starts_with("list2_"), starts_with("list3_"), starts_with("list4_"))) %>% 
  mutate(source = str_sub(name, 1, -3), list_yr = as.numeric(paste0("20", str_sub(name, -2)))) %>% 
  rename(seen_in_source = value) %>% 
  mutate(
    age = list_yr - byear,
    age_cat = if_else(
      list_yr - byear <= 18, "13-18", "19-25"
    )) %>% 
  select(sex, byear, race_cat, age, age_cat, list_yr, source, seen_in_source, id)  


dat_long2 <- dat_long %>%
  left_join(dat_elig) %>% 
  filter(seen_in_source == 1 & eligible == 1) %>% # keep only person-year records in which an individual was observed while within the study age range
  select(-eligible) %>% 
  distinct()

reactable(dat_long2 %>% slice_sample(n = 10))

```



Next, we split the dataset into strata of relatively homogenous subgroups of youth, defined by every unique combination of sex, race, age group, and year (e.g., sex == MALE, race = NONWHITE, age group = 13-18, list year == 2017). 



```{r}
make_strata <- dat_long2 %>% 
  group_by(sex, race_cat, age_cat, list_yr) %>% 
  summarise(
    n_kids = n_distinct(id)
  ) %>% 
  ungroup() %>% 
  mutate(strata = row_number())

dat_long3 <- make_strata %>% 
  left_join(dat_long2)

```


We then reshape our dataset again, from "long" to "wide" to prepare it for the analysis step. 


```{r}
dat_wide <- dat_long3 %>% 
  pivot_wider(names_from = source, values_from = seen_in_source, values_fill = 0)

reactable(dat_wide %>% slice_sample(n = 10))

```


The final data-reshaping step is to turn the person-level dataset into an aggregated strata-level dataset needed for estimating population sizes with MSE. First, we create a 'split' data frame with 24 rows- one for each stratum. This is accomplished with the base::split function, which groups observations according to the specified variable (strata in this case) and collapses the subset of records within each subgroup into a new column named _value_ by default. In other words, whereas tables are typically organized to include a single piece of information per cell, each cell in _value_ is an entire dataset with multiple columns.

Structuring the records in this way comes in handy when the same set of analyses are to be performed on many subsets of the data. For one, the code only needs to be written once and run once, as it is applied simultaneously to each subset of the data. In other words, it eliminates the need for tedious (and error-prone) copying, pasting, and re-running to run the code on each subgroup. Second, it saves the output of each analysis step as a new column or set of columns in the dataframe, making it easy to directly compare the results for each subgroup.  

```{r}
by_strata <- dat_wide %>% 
  split(.$strata) %>% 
  enframe()

vars <- dat_wide %>% 
  select(strata, sex, age_cat,list_yr, race_cat, n_kids) %>% 
  distinct()

by_strata <- vars %>% 
  bind_cols(by_strata)


```

The function _make_matrix_, defined below, takes our split dataset and the strata id (named 1:24) as inputs. For each stratum, it drops any observation list in which no individuals were observed, and create a matrix of the form needed to run the MSE analysis. 

```{r}

make_matrix <- function(dw, strata){

# Create a dataframe showing the combination of list observations for each record. The variable lp indicates the list pattern for each individual, based on the column order of the lists.

d2<- dw %>% 
  unite(lp, -c(sex:id), remove = FALSE) %>% 
  select( lp:ncol(.)) 


# For each subgroup, count the number of unique contributions made by each list (i.e., the number of individuals observed in HMIS alone, and not in combination with any other list.). Drop any lists that do not make at least one unique/independent contribution

d2$nlists <- rowSums(d2[,-1])




todrop <- d2 %>% 
  filter(nlists == 1) %>% 
  summarise(across(-c(lp, nlists), sum)) %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  filter(V1 == 0)

todrop <- todrop$rowname



# remove any zero-contribution lists from the dataset, re-create the variable lp to show the observation patterns of only the included lists.
d2 <- d2 %>% 
  select(-c(all_of(todrop), nlists)) %>% 
  select(-lp) %>% 
  unite(lp, remove = FALSE)



d3 <- d2 %>% 
  group_by( lp) %>% 
  tally() %>% 
  ungroup() 



dt <- merge(d2, d3) 
d0<- distinct(dt, lp, .keep_all = TRUE) %>% 
  mutate(across(everything(), \(x) replace_na(x, 0)))

d00 <- d0 %>% 
  select(-lp) 

d00

}


```

In the following step, we begin adding new columns to the split dataset. Specifically, the variable _mat_ contains the observation matrix for each stratum, while _vars_ shows the homelessness lists in which individuals within that stratum were observed at least once. Depending on the population of interest and the available homeless indicator databases, _vars_ may or may not be relevant. In the example data, we can see that there were no individuals observed in the list4_ (juvenile delinquency & department of child & family services databases) databases in several of the age 19-25 strata. Thus, this list does not appear in the list history matrices for these strata.

The observation matrix for stratum 1, Non-white females, aged 13-18 in 2017, is shown below

```{r}
by_strata2 <- by_strata %>%  
  mutate(mat = map(value, ~make_matrix(.x))) %>% 
  mutate(vars = map(mat, names)) %>% 
  select(-c(value))

reactable(by_strata2[[8]][[1]])

```



The function _mse_fun_ applies multiple functions from the SparseMSE package to each stratum, and adds the output of those functions as new columns in the split dataframe. For more details on these functions, please refer to the journal article and R documentation referenced at the beginning of this document. 

checkident performs a test to determine whether it is possible to generate a consistent estimate of population size given the observation history for a given set of observations. A value of 0 that the conditions for consistent estimation are met, while values of 1, 2, and 3 indicate issues of MLE existence, identifiability, or both. 

The estimatepopulation.0 function estimates the population for each stratum, including the 'hidden figure' of homeless youth who are not observed on any list. The output of this function is saved as the variable est_stepwise. 

The variable est_pt_sw is the point estimate of the total population, including both the observed and unobserved (hidden) cases. We use this value to construct additional useful variables, such as the number of unobserved youth, and the ratio of unobserved:observed youth. 

```{r}

mse_fun <- function(df) {
by_strata2 <- df %>%  
  mutate(checkident = map_dbl(mat, ~checkident(.x, mX=0,verbose=FALSE)))


by_strata3 <- by_strata2 %>% 
  mutate(est_stepwise = map(mat, ~estimatepopulation.0(.x,  quantiles=c(0.025,0.975)))) %>%
  mutate(
    est_low_sw = map_dbl(est_stepwise, ~pluck(.x, "estimate", "2.5%" )),
    est_pt_sw = map_dbl(est_stepwise, ~pluck(.x, "estimate", "point est." )),
    est_hi_sw = map_dbl(est_stepwise, ~pluck(.x, "estimate", "97.5%" ))) 

test <- by_strata3 %>% 
  mutate(across(est_low_sw:est_hi_sw, round), 
         est_unlisted = est_pt_sw - n_kids, 
         unlisted_to_listed = round(est_unlisted/n_kids, 1)) %>% 
  select(sex, age_cat, list_yr, race_cat, n_kids, est_pt_sw, est_low_sw, est_hi_sw, est_unlisted, unlisted_to_listed, checkident)


  
}

```



```{r message=FALSE, warning=FALSE}
by_mse <- mse_fun(by_strata2)

reactable(by_mse)
```

